---
title: "Medicare Association Rule Mining & Audit Flagging"
author: "Christopher Reddish"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

required_packages <- c("dplyr", "readr", "arules", "ggplot2", "ggrepel","purrr", "tibble")
to_install <- setdiff(required_packages, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
lapply(required_packages, require, character.only = TRUE)
```

## 📘 Introduction

This project detects potentially suspicious billing behavior among Medicare providers in the Tampa Bay area. Using association rule mining and a composite scoring model, it flags providers who may warrant audit review.

---

## 🔹 Step 1: Clean and Prepare Data

```{r step-1-clean-data}
filtered_data <- read.csv("filtered_data.csv")

clean_df <- filtered_data %>%
  filter(
    !is.na(Provider_Type),
    !is.na(HCPCS_Code),
    !is.na(NPI),
    as.numeric(Avg_Medicare_Payment) >= 10
  ) %>%
  mutate(
    NPI = as.character(NPI),
    HCPCS_Code = as.character(HCPCS_Code),
    Provider_Type = as.factor(trimws(Provider_Type))
  ) %>%
  select(NPI, HCPCS_Code, Avg_Medicare_Payment, Provider_Type, Provider_Last_Name)

```

## 🔹 Step 2: Generate Association Rules by Specialty


```{r step-2-rules-mining, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Initialize lists to store results
all_rule_matches <- tibble()      # For NPI-rule match scoring in Step 3
provider_summary_list <- list()   # For rule-based provider summaries

# Step 2A: Split the dataset by specialty (Provider_Type)
specialty_groups <- clean_df %>%
  group_by(Provider_Type) %>%
  group_split()

# Step 2B: Loop through each specialty subset
for (spec_df in specialty_groups) {
  spec_name <- as.character(unique(spec_df$Provider_Type))
  if (nrow(spec_df) < 2) next

  # Step 2C: Convert each NPI into a transaction of billed HCPCS codes
  provider_tx <- spec_df %>%
    group_by(NPI) %>%
    summarise(items = list(unique(HCPCS_Code)), .groups = "drop")

  tx_list <- lapply(split(provider_tx$items, provider_tx$NPI), unlist)
  if (length(tx_list) < 2) next

  trans <- as(tx_list, "transactions")

  # Step 2D: Generate association rules for this specialty
  rules <- apriori(trans, parameter = list(supp = 0.02, conf = 0.7, maxlen = 2))
  if (length(rules) == 0) next

  # Step 2E: Filter rules by lift and convert to data frame
  rules_df_raw <- as(rules, "data.frame")
  rules_df_filtered <- rules_df_raw %>% filter(lift >= 2)
  if (nrow(rules_df_filtered) == 0) next

  # Add readable labels and parse rule structure for each rule
  rules_df_filtered$lhs <- labels(lhs(rules))[as.integer(rownames(rules_df_filtered))]
  rules_df_filtered$rhs <- labels(rhs(rules))[as.integer(rownames(rules_df_filtered))]
  rules_df_filtered <- rules_df_filtered %>%
    mutate(
      rule_label = paste(lhs, "=>", rhs),
      lhs_items = strsplit(gsub("[{}]", "", lhs), ","),
      rhs_items = strsplit(gsub("[{}]", "", rhs), ","),
      all_items = map2(lhs_items, rhs_items, ~ unique(trimws(c(.x, .y))))
    )

  # Step 2F: Match rules to providers
  provider_rule_map <- list()
  npi_rule_log <- tibble()

  for (npi in names(tx_list)) {
    provider_items <- tx_list[[npi]]

    matched <- rules_df_filtered %>%
      filter(map_lgl(all_items, ~ all(.x %in% provider_items)))

    if (nrow(matched) > 0) {
      provider_rule_map[[npi]] <- matched$rule_label

      npi_rule_log <- bind_rows(npi_rule_log, tibble(
        NPI = npi,
        Rule_Label = matched$rule_label,
        Lift = matched$lift,
        Confidence = matched$confidence
      ))
    }
  }

  # Step 2G: Create provider summary stats for rule match counts
  if (length(provider_rule_map) > 0) {
    provider_summary <- tibble(
      Specialty = spec_name,
      NPI = names(provider_rule_map),
      Total_Rules = sapply(provider_rule_map, function(x) length(unique(x)))
    )

    hcpcs_diversity <- spec_df %>%
      group_by(NPI) %>%
      summarise(Unique_Codes = n_distinct(HCPCS_Code), .groups = "drop")

    provider_summary <- provider_summary %>%
      left_join(hcpcs_diversity, by = "NPI") %>%
      mutate(
        Rules_Per_Code = round(Total_Rules / Unique_Codes, 2),
        Unique_Codes_Per_Rule = round(Unique_Codes / Total_Rules, 4)
      )

    provider_summary_list[[spec_name]] <- provider_summary
    all_rule_matches <- bind_rows(all_rule_matches, npi_rule_log)
  }
}

# Export objects for Step 3
final_provider_results <- bind_rows(provider_summary_list)
npi_rule_scores <- all_rule_matches %>%
  group_by(NPI) %>%
  summarise(
    mean_lift = mean(Lift, na.rm = TRUE),
    mean_confidence = mean(Confidence, na.rm = TRUE),
    .groups = "drop"
  )

```

## 🔹 Step 3:  Step 3A: Merge all provider summaries across specialties into one dataset

```{r step-3-scoring}
# Step 3A: Merge rule-based stats into the full cleaned Medicare dataset
merged_df <- clean_df %>%
  left_join(final_provider_results, by = "NPI")

# Step 3B: Append average lift and confidence per provider from rule matches
scored_providers <- merged_df %>%
  left_join(npi_rule_scores, by = "NPI")

# Step 3C: Z-score standardization for Rules_Per_Code within each specialty
scored_providers <- scored_providers %>%
  group_by(Provider_Type) %>%
  mutate(Z_Score_RPC = as.numeric(scale(Rules_Per_Code))) %>%
  ungroup()

# Step 3D: Compute a composite audit score
# This blends:
#   - billing repetition
#   - rule lift
#   - rule confidence
#   - average Medicare payout
#   - relative specialty behavior
scored_providers <- scored_providers %>%
  mutate(
    Composite_Audit_Score = round(
      (coalesce(Rules_Per_Code, 0) * 1.5) +
      (coalesce(mean_lift, 0) * 2.0) +
      (coalesce(mean_confidence, 0) * 1.0) +
      (coalesce(Avg_Medicare_Payment, 0) / 100) +
      coalesce(Z_Score_RPC, 0),
      2
    )
  )

# Step 3E: Append sample size per provider for audit traceability
scored_providers <- scored_providers %>%
  group_by(NPI) %>%
  mutate(Sample_Size = n()) %>%
  ungroup()

```

## 🔹 Step 4: Risk Stratification & Visualization

```{r step-4-risk-visualization, fig.width=18, fig.height=11}
# Step 4A: Flag audit risk levels using percentile thresholds
# ------------------------------------------------------------
# We classify providers within each specialty based on Composite Score:
# - Top 2% = Extreme High Risk
# - 92–98% = High Risk
# - 80–92% = Moderate Risk
# - Bottom 80% = Low/No Risk

scored_flagged <- scored_providers %>%
  group_by(Provider_Type) %>%
  mutate(
    p80 = quantile(Composite_Audit_Score, 0.80, na.rm = TRUE),
    p92 = quantile(Composite_Audit_Score, 0.92, na.rm = TRUE),
    p98 = quantile(Composite_Audit_Score, 0.98, na.rm = TRUE),
    Audit_Flag = case_when(
      Composite_Audit_Score >= p98 ~ "Extreme High Risk (98–100%)",
      Composite_Audit_Score >= p92 ~ "High Risk (92–98%)",
      Composite_Audit_Score >= p80 ~ "Moderate Risk (80–92%)",
      TRUE                         ~ "Low/No Risk (0–80%)"
    )
  ) %>%
  ungroup()

# Step 4B: Filter to valid (non-NA, >0) composite scores
valid_scores_df <- scored_flagged %>%
  filter(!is.na(Composite_Audit_Score), Composite_Audit_Score > 0)

# Step 4C: Manually set your "Top 15 Most Interesting" specialties
top15_specialties <- c(
  "Anesthesiology", "Cardiology", "Dermatology", "Diagnostic Radiology",
  "Emergency Medicine", "Family Practice", "Gastroenterology", "General Surgery",
  "Internal Medicine", "Interventional Cardiology", "Nurse Practitioner",
  "Orthopedic Surgery", "Physician Assistant", "Podiatry", "Vascular Surgery"
)

# Step 4D: Filter only the specialties of interest and cap extreme outliers
volume_filtered_df <- valid_scores_df %>%
  filter(
    Provider_Type %in% top15_specialties,
    Composite_Audit_Score <= quantile(Composite_Audit_Score, 0.99, na.rm = TRUE),
    Avg_Medicare_Payment <= 1000
  )

# Step 4E: Generate plot with fixed x and y axis, no log scale
composite_plot <- ggplot(volume_filtered_df, aes(
  x = Avg_Medicare_Payment,
  y = Composite_Audit_Score,
  color = Audit_Flag
)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Provider_Type, scales = "fixed", nrow = 5) +
  scale_color_manual(
    values = c(
      "Extreme High Risk (98–100%)" = "purple",
      "High Risk (92–98%)" = "red",
      "Moderate Risk (80–92%)" = "gold",
      "Low/No Risk (0–80%)" = "green"
    )
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    breaks = c(0, 250, 500, 750, 1000),
    limits = c(0, 1000)
  ) +
  scale_y_continuous(
    limits = c(0, 85),
    breaks = seq(0, 80, by = 20)
  ) +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(size = 10)) +
  labs(
    title = "Composite Audit Score vs. Medicare Payment",
    subtitle = "Top 15 Most Interesting Specialties – Risk Segmentation at 80/92/98 Percentiles",
    x = "Average Medicare Payment ($)",
    y = "Composite Audit Score",
    color = "Audit Risk Level"
  )

# Step 4F: Save output as PNG
ggsave("Top15_Composite_vs_Payment_BySpecialty_FIXEDAXIS.png",
       plot = composite_plot, dpi = 300, width = 14, height = 9)

# Step 4G: Render in output
composite_plot

```

## 🔹 Step 5: Top Providers Report

```{r step-5-top-providers}
# Step 5A: Join Provider_Last_Name from filtered_data, and rename it cleanly
scored_named <- scored_providers %>%
  left_join(
    filtered_data %>%
      mutate(NPI = as.character(NPI)) %>%
      select(NPI, Provider_Last_Name),
    by = "NPI"
  ) %>%
  rename(Provider_Last_Name = Provider_Last_Name.y)  # in case it became .y

# Step 5B: Top 50 providers overall
top50_overall <- scored_named %>%
  distinct(NPI, .keep_all = TRUE) %>%
  arrange(desc(Composite_Audit_Score)) %>%
  slice_head(n = 50)

# Display Top 50 Table
knitr::kable(
  top50_overall %>%
    select(Provider_Last_Name, NPI, Provider_Type, Composite_Audit_Score),
  caption = "Top 50 Providers Overall by Composite Audit Score"
)

# Step 5C: Get Top 3 Providers by Specialty without repeating NPIs
top3_by_specialty <- scored_named %>%
  filter(!is.na(Composite_Audit_Score)) %>%
  group_by(Provider_Type, NPI) %>%
  slice_max(order_by = Composite_Audit_Score, n = 1) %>%  # get max score per NPI
  ungroup() %>%
  distinct(NPI, .keep_all = TRUE) %>%                     # remove any duplicate NPIs
  group_by(Provider_Type) %>%
  slice_max(order_by = Composite_Audit_Score, n = 3) %>%  # now take top 3 per specialty
  ungroup()

# Display Top 3 Per Specialty Table
knitr::kable(
  top3_by_specialty %>%
    select(Provider_Type, Provider_Last_Name, NPI, Composite_Audit_Score),
  caption = "Top 3 Providers by Composite Score in Each Specialty"
)

```
