---
title: "CMS Final Project"
author: Christopher Reddish, Khushi Patel, Naomi Vaid, Sneh Patel, Scott Young,  and
  Steven Barden
date: "04/20/2025"
output: html_document
---

## Introduction

```{r, "Libraries"}
# Add your library below.
# List of required packages
required_packages <- c(
  "tidyverse",
  "GGally",
  "ggplot2",
  "sf",
  "dplyr",
  "readr",
  "arules",
  "leaflet",
  "tigris",
  "scales",
  "htmlwidgets",
  "jsonlite"
)

# Install any packages that are not already installed
to_install <- setdiff(required_packages, rownames(installed.packages()))
if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}

# Load all required packages and check if successfully loaded
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    warning(sprintf("Package '%s' could not be loaded.", pkg))
  }
}

# Enable caching for tigris package
options(tigris_use_cache = TRUE)

```

##Cleaning the dataset: Sneh Patel, Steven Barden, Christopher Reddish
1. Filtering out Florida Data
2. Adding Custom Labels for column
3. Removing unwanted columns 
```{r, "Step 1"}
## Step 1.1
# Read CSV 
filtered_data <- read_csv("Data/Medicare_dataset.csv", show_col_types = FALSE) %>%
  mutate(Rndrng_Prvdr_Zip5 = ifelse(Rndrng_Prvdr_Zip5 %in% c("19107", "19144"), "33308", Rndrng_Prvdr_Zip5)) %>%
  filter(grepl("^\\d+$", Rndrng_Prvdr_Zip5)) %>%
  mutate(Rndrng_Prvdr_Zip5 = as.numeric(Rndrng_Prvdr_Zip5)) %>%
  filter(Rndrng_Prvdr_Zip5 >= 32001 & Rndrng_Prvdr_Zip5 <= 34999) %>%
  # Rename columns
  rename(
    NPI = Rndrng_NPI,
    Provider_Last_Name = Rndrng_Prvdr_Last_Org_Name,
    Provider_First_Name = Rndrng_Prvdr_First_Name,
    Provider_MI = Rndrng_Prvdr_MI,
    Provider_Credentials = Rndrng_Prvdr_Crdntls,
    Entity_Code = Rndrng_Prvdr_Ent_Cd,
    Provider_Street1 = Rndrng_Prvdr_St1,
    Provider_Street2 = Rndrng_Prvdr_St2,
    Provider_City = Rndrng_Prvdr_City,
    Provider_State = Rndrng_Prvdr_State_Abrvtn,
    FIPS_Code = Rndrng_Prvdr_State_FIPS,
    Provider_Zip = Rndrng_Prvdr_Zip5,
    RUCA_Code = Rndrng_Prvdr_RUCA,
    RUCA_Description = Rndrng_Prvdr_RUCA_Desc,
    Provider_Country = Rndrng_Prvdr_Cntry,
    Provider_Type = Rndrng_Prvdr_Type,
    Medicare_Participation = Rndrng_Prvdr_Mdcr_Prtcptg_Ind,
    HCPCS_Code = HCPCS_Cd,
    HCPCS_Description = HCPCS_Desc,
    HCPCS_Drug_Indicator = HCPCS_Drug_Ind,
    Place_Of_Service = Place_Of_Srvc,
    Total_Beneficiaries = Tot_Benes,
    Total_Services = Tot_Srvcs,
    Total_Beneficiary_Service_Days = Tot_Bene_Day_Srvcs,
    Avg_Submitted_Charge = Avg_Sbmtd_Chrg,
    Avg_Medicare_Allowed = Avg_Mdcr_Alowd_Amt,
    Avg_Medicare_Payment = Avg_Mdcr_Pymt_Amt,
    Avg_Medicare_Standardized_Amt = Avg_Mdcr_Stdzd_Amt
  ) %>%
  # Drop unnecessary columns
  select(-Provider_First_Name, -Provider_MI)

# Save the filtered data to a new CSV in the 'data' folder
write.csv(filtered_data, "Data/filtered_data.csv", row.names = FALSE)

# Print confirmation
print(paste("Filtered data saved to: Data/filtered_data.csv"))

```
## Load Data Once and Custom Functions
```{r, "Load Data"}
#Load the filtered data ONCE after cleaning and take care of problems of FIPS code error and zip map for k-mean map
filtered_data <- read_csv("Data/filtered_data.csv",
                          col_types = cols(FIPS_Code = col_character())) %>%
  mutate(FIPS_Code = as.numeric(gsub("[^0-9]", "", FIPS_Code)), Provider_Zip = sprintf("%05s", as.character(Provider_Zip)))


# Helper function to clean monetary columns
clean_money <- function(x) {
  as.numeric(gsub("[$,]", "", as.character(x)))
}

```

## Association Rule Mining: Christopher Reddish, Scott Young
 
```{r, "Step 2"}
transactions_data <- filtered_data %>%
  group_by(Provider_Type, NPI) %>%
  summarize(HCPCS_Description = list(unique(HCPCS_Description)), .groups = "drop") %>%
  ungroup() %>%
  select(HCPCS_Description) %>%
  pull()
transactions <- as(transactions_data, "transactions")

rules <- apriori(transactions, parameter = list(supp = 0.01, conf = 0.5, minlen = 2))
inspect(head(sort(rules, by = "lift"), 10))
rules_df <- as(rules, "data.frame")
```


## K-Means Clustering: Christopher Reddish, Sneh Patel
# The primary goal of this analysis is to analyze Medicare provider data using k-means clustering, assign meaningful cluster labels to providers based on cost and service volume, and visualize the geographic distribution of these clusters across ZIP codes in Florida using an interactive map..

```{r, "Step 3"}
# ---- Step 3: K-Means Clustering ----
# K-Means Clustering
df_raw <- filtered_data %>%
  mutate(
    Total_Services = as.numeric(Total_Services),
    Avg_Submitted_Charge = clean_money(Avg_Submitted_Charge),
    Avg_Medicare_Payment = clean_money(Avg_Medicare_Payment)
  ) %>%
  select(Total_Services, Avg_Submitted_Charge, Avg_Medicare_Payment) %>%
  na.omit()

stopifnot(all(sapply(df_raw, is.numeric)))

df_scaled <- scale(df_raw)
k <- min(4, nrow(unique(df_scaled)))
if (k < 2) stop("Not enough data to form clusters.")

set.seed(123)
clusters <- kmeans(df_scaled, centers = k, nstart = 25)

cluster_labels <- c(
  "1" = "Moderate Cost, High Volume",
  "2" = "Typical Providers",
  "3" = "High Cost, Low Volume",
  "4" = "Low Cost, Very High Volume"
)

results <- df_raw %>%
  mutate(
    Cluster = clusters$cluster,
    Cluster_Label = cluster_labels[as.character(Cluster)]
  )

# Mapping
zcta_shapes <- tigris::zctas(cb = TRUE, year = 2020) %>%
  rename(Zip = ZCTA5CE20) %>%
  mutate(Zip = sprintf("%05s", as.character(Zip)))

zip_cluster_data <- results %>%
  mutate(Provider_Zip = filtered_data$Provider_Zip) %>%
  group_by(Provider_Zip, Cluster_Label) %>%
  summarise(Count = n(), .groups = "drop")

zip_cluster_sf <- left_join(zcta_shapes, zip_cluster_data, by = c("Zip" = "Provider_Zip")) %>%
  filter(!is.na(Cluster_Label)) %>%
  st_make_valid() %>%
  filter(st_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>%
  st_transform(4326)

zip_cluster_sf <- zip_cluster_sf %>%
  mutate(
    fill_color = dplyr::recode(Cluster_Label,
      "Moderate Cost, High Volume" = "#1f78b4",
      "Typical Providers" = "#33a02c",
      "High Cost, Low Volume" = "#e31a1c",
      "Low Cost, Very High Volume" = "#ff7f00",
      .default = "#cccccc"
    ),
    opacity = scales::rescale(Count, to = c(0.4, 1))
  )

# Interactive Leaflet Map
leaflet_map <- leaflet::leaflet(zip_cluster_sf) %>%
  leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron) %>%
  leaflet::addPolygons(
    fillColor = ~fill_color,
    fillOpacity = ~opacity,
    color = "#444444",
    weight = 0.5,
    popup = ~paste0(
      "<b>ZIP:</b> ", Zip, "<br>",
      "<b>Cluster:</b> ", Cluster_Label, "<br>",
      "<b>Provider Count:</b> ", Count
    )
  ) %>%
  leaflet::addLegend(
    "bottomright",
    colors = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00"),
    labels = c(
      "Moderate Cost, High Volume",
      "Typical Providers",
      "High Cost, Low Volume",
      "Low Cost, Very High Volume"
    ),
    title = "Cluster Type",
    opacity = 1
  )

htmlwidgets::saveWidget(leaflet_map, "Medicare_Clusters_FL.html", selfcontained = TRUE)

```
# Summary of Findings
# According to the provider statistics, Cluster 2 is the most common,
# accounting for the majority of providers, while Cluster 3 is under-represented, implying potential clustering imbalances. 
# Cluster 2 has a wide range of values for total services, average submitted charge, and average Medicare payment, 
# with average Medicare payments at $70.23, average submitted charges at $356.84, and an average of 604 total services; 
# however, Cluster 3 has higher costs and lower service volumes, with average Medicare payments at $425.36, 
# average submitted charges at $4,490.86, and an average of 29 total services.
# These findings indicate that the clustering technique should be reconsidered, possibly using new algorithms or feature engineering, 
# as well as a more rigorous outlier analysis within Cluster 2 to assess whether some providers should be reclassified in order to establish more distinct and balanced groups.

## Linear Regression: Steven Barden, Scott Young
 
```{r, "Step 4"}
## Linear Regression (Step 4)
# Load and clean with basic numeric conversion
df_lm <- filtered_data %>%
  select(
    Avg_Medicare_Payment,
    Total_Services,
    Avg_Submitted_Charge,
    Total_Beneficiaries,
    Place_Of_Service,
    Provider_Type,
    RUCA_Description
  ) %>%
  mutate(
    Medicare_Payment = parse_number(as.character(Avg_Medicare_Payment)),
    Services = parse_number(as.character(Total_Services)),
    Submitted_Charge = parse_number(as.character(Avg_Submitted_Charge)),
    Beneficiaries = parse_number(as.character(Total_Beneficiaries)),
    Place_Of_Service = factor(Place_Of_Service),
    Specialty = factor(Provider_Type),
    RUCA_Description = factor(RUCA_Description)
  ) %>%
  select(-Avg_Medicare_Payment, -Total_Services, -Avg_Submitted_Charge, -Total_Beneficiaries, -Provider_Type) %>%
  drop_na()

# Run simplified regression (removed HCPCS_Description)
lm_model <- lm(
  Medicare_Payment ~ Services + Submitted_Charge + Beneficiaries + 
    Place_Of_Service + Specialty + RUCA_Description,
  data = df_lm
)

# View results
summary(lm_model)
```
# The linear model uses Services, Submitted_Charge, and Beneficiaries as the numeric predictors.
# The categorical predictors are Place_Of_Srvc, Specialty, and RUCA_Desc.
# The R-squared value of the model is 0.6486, which explains a 65%~ variation in Medicare_Payment.
# The F-statistic & P-values are both quite high, indicating a statistically significant model.
# The average error stands at about $176.
# As the Services increase, the average Medicare_Payment slightly decreases which tells that dense facilities reflect an economy-of-scale factor.
# The strongest predictor is the Submitted_Charge value, in which higher billed values raises Medicare payments, as is obvious.
# Beneficiaries stands as not significant, meaning the amount of people receiving service does not influence costs.
# Place_Of_Srvc0, in an office setting, has a high p-value of $20.20. This stands to reason that smaller facilities are associated with a higher cost.
# In terms of other categorical variables and their effect, Specialty has a range of types in which costs are reduced or increased on average.
## Increases include: Ambulance Service Provider, Oral Surgery, Peripheral Vascular Disease, and Micrographic Dermatologic Surgery.
## Decreases include: CRN and Anesthesiology.
# RUCA_Desc has the following takeaways: Micropolitan areas receive less payment, Secondary flow to large urbanized areas has a higher payment.

## Heat Map Visual: Khushi Patel, Naomi Vaid
 
```{r, "Step 5"}

# File paths
shapefile_path <- "~/Data/tl_2010_12_zcta510.shp"

# Load Medicare data
medicare_cleaned <- filtered_data %>%
  mutate(
    Provider_Zip = as.character(Provider_Zip),
    Avg_Medicare_Payment = as.numeric(gsub("[$,]", "", Avg_Medicare_Payment))
  )

# Load shapefile
zcta <- st_read(shapefile_path) %>%
  mutate(ZCTA5CE10 = as.character(ZCTA5CE10))

# Choose specialty
selected_specialty <- "Internal Medicine"

# Filter and summarize
filtered_data <- medicare_cleaned %>%
  filter(Specialty == selected_specialty) %>%
  group_by(Provider_Zip) %>%
  summarize(
    Avg_Payment = mean(Avg_Medicare_Payment, na.rm = TRUE),
    Providers = n(),
    .groups = "drop"
  )

# Merge with shapefile
merged_data <- zcta %>%
  left_join(filtered_data, by = c("ZCTA5CE10" = "Provider_Zip"))

# Create heat map
ggplot(data = merged_data) +
  geom_sf(aes(fill = Avg_Payment), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey90",
    name = "Avg Payment ($)",
    limits = c(0, 150),
    breaks = c(0, 50, 100, 150),
    oob = scales::squish
  ) +
  labs(
    title = "Medicare Average Payment by ZIP Code in Florida",
    subtitle = paste("Specialty:", selected_specialty),
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),
    plot.caption = element_text(size = 10, face = "italic")
  )

# Here is a summary of our analysis based on Medicare payment data visualization:

# We analyzed average Medicare payments for providers specializing in Internal Medicine across Florida ZIP codes. 
# This process began by loading and preparing a cleaned Medicare dataset and a shapefile representing Florida ZIP Code Tabulation Areas (ZCTAs).

# After filtering the data for the selected specialty ("Internal Medicine"), we calculated the average Medicare 
# payment and the number of providers per ZIP code. This summarized dataset was then merged with the shapefile data 
# to enable geographic plotting.

# Using ggplot2 and the sf package, we created a choropleth map showing the distribution of average Medicare payments 
# across Florida. Areas with higher average payments appear in brighter plasma shades, while areas with lower payments 
# or no data are shown in muted tones.

# The heatmap helps reveal regional disparities in Medicare reimbursements, highlighting ZIP codes with notably higher 
# or lower average payments. These visual insights can inform further investigation into geographic trends in healthcare 
# spending, potential inequities in reimbursement rates, or gaps in provider coverage for Internal Medicine in Florida.
```
