---
title: "CMS Final Project"
author: Christopher Reddish, Khushi Patel, Naomi Vaid, Sneh Patel, Scott Young,  and
  Steven Barden
date: ""
output: html_document
---

## Introduction

```{r, "Libraries"}
# Add your library below.


```

##Cleaning the dataset: Sneh Patel, Steven Barden, Christopher Reddish
1. Filtering out Florida Data
2. Adding Custom Labels for column (needed)
3. Removing unwanted columns (needed)
```{r, "Step 1"}
## Step 1.1
# Read CSV 
# data <- read.csv("data/OGdataCMS.csv")  

# Filter rows where zipcode is between 32001 and 34999
# filtered_data <- data[data$Rndrng_Prvdr_Zip5 >= 32001 & data$Rndrng_Prvdr_Zip5 <= 34999, ]

# Save the filtered data to a new CSV in the 'data' folder
# write.csv(filtered_data, "data/filtered_data.csv", row.names = FALSE)

# Print confirmation
print(paste("Filtered data saved to: data/filtered_data.csv"))
```

## Association Rule Mining: Christopher Reddish, Scott Young
 
```{r, "Step 2"}

```


## K-Means Clustering: Christopher Reddish, Sneh Patel

```{r, "Step 3"}

```


## Linear Regression: Steven Barden, Scott Young
 
```{r, "Step 4"}

```



## Heat Map Visual: Khushi Patel, Naomi Vaid
 
```{r, "Step 5"}

```
